---
title: "Comscore Data Cleaning"
author: "Megan McQueen"
date: "1/14/2022"
output: html_document
---

Packages: 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# data cleaning
library(dplyr)
library(ggplot2)
library(magrittr)
# for heatmaps
library(ComplexHeatmap)
library(pheatmap)
library(RColorBrewer)
# cluster analysis and autoencoding
library(h2o)
library(cluster)
library(factoextra)
# date/time processing
library(lubridate)
library(hms)
```

## Read in data

Save as RDS file for faster access later.
```{r}
df0 <- read.csv("/Users/sebastiantsai/Desktop/Applied Math Thesis/raw data/browsing_sample_2017.csv")
saveRDS(df0, file = "/Users/sebastiantsai/Desktop/Applied Math Thesis/raw data/browsing_sample_2017.rds")
df0 <- readRDS(file = "/Users/sebastiantsai/Desktop/Applied Math Thesis/raw data/browsing_sample_2017.rds")

```

## Subset IDs

Select subset of households from original dataset for test analysis.

```{r}
# how many houses originally?
nhouseholds <- length(unique(df0$machine_id))
print(paste(nhouseholds, "households total"))

# sample n households
n <- 300
set.seed(1)
samp_ids <- sample(unique(df0$machine_id),n)
print(paste("sampling",n,"household(s)"))
# include only above households in dataset
df <- df0[df0$machine_id %in% samp_ids,]
# new dataset length (rows)
print(paste(nrow(df), "observations for",n,"household(s)"))
# confirm no. households is correct
#print(paste(length(unique(df$machine_id)) == n, "- no. households makes sense"))
```

## Date formatting

### Function: Add date/time columns to dataframe
```{r lubridate-try1}
cleantimes <- function(df) {
  # parse dates
  df$event_date.ld <- ymd(df$event_date)
  # weekday of session start
  df$start_wday <- wday(df$event_date.ld, label = FALSE)
  # parse times
  df$event_time.ld <- hms::as_hms(df$event_time)

  # length of session (period - minutes)
  df$duration.ld <- minutes(df$duration)
  # get date-time of session start
  df$event_datetime <- ymd_hms(paste(as.character(df$event_date.ld), as.character(df$event_time.ld)))
  # get session end date/time using duration and start date
  df$event_end <- df$event_datetime + df$duration.ld
  # weekday of session end
  df$end_wday <- wday(df$event_end)
  # start and end hours
  df$start_hour <- hour(df$event_time.ld)
  df$end_hour <- hour(df$event_end)
  
  return(df)
}


unique(df0.time$start_hour)
temp <- df0.time %>% 
  select(machine_id, event_date, event_time, duration)

df0.time <- cleantimes(df0)
#df0.time <- readRDS(file = "/Users/meganmcqueen/Box/Senior Project/data/browsing_sample_2017_cleanedTimes.rds")
df0.time
```

Read in data after cleaning dates and times:

```{r}
df0.time <- readRDS(file = "/Users/meganmcqueen/Box/Senior Project/data/browsing_sample_2017_cleanedTimes.rds")
paste0("Data from ",min(df0.time$event_datetime), " to ", max(df0.time$event_datetime))
```

Look at session lengths/hours on average:

```{r time-graphs-1, message=F}
# what proportion of sessions less than 1 hour?
mean(df$duration < 60)
# avg duration
mean(df$duration)
# plot durations
df %>% ggplot(aes(duration)) + geom_histogram() + ggtitle("Session Durations in Minutes")
```

## Single matrix data structure

+ Number of columns = 168 (7*24) + 1 = 169 
+ Number of rows = number of households

Let each day of the week (Sun - Sat) correspond with a value 1-7, $d$. 

+ Sun = 1
+ Mon = 2
+ Tues = 3
+ Wed = 4
+ Thurs = 5
+ Fri = 6
+ Sat = 7

And let the hour of the day, from $0$ to $23$, be denoted by $h$. 

In column 1 we have the data for hour 0 of day 1 (Sunday). 

For a given day of week $d$, the columns corresponding to it are $(24d-23):24d$. 
The column corresponding to hour $h$ on day $d$ is number $24d + h - 23$. 

The last column, col $169$, contains the machine ID. 


```{r redo-household-vector}
# Function takes in a dataframe of internet use data
# And outputs a vector with 24*7 entries containing activity stats
household.vec <- function(df) {
  # empty 7*24 VECTOR
  m <- rep(0,24*7)
  
  # break up df into observations where session remained w/in same day
  df.sameday <- df %>% filter(start_wday == end_wday)
  # populate matrix for observations occurring within same day
  for (i in 1:nrow(df.sameday)) {
    day <- df.sameday$start_wday[i]
    hour1 <- df.sameday$start_hour[i] 
    hour2 <- df.sameday$end_hour[i]
    # increment entry by 1 b/c observed activity during that hour/day of week
    h1.index <- 24*day + hour1 - 23
    h2.index <- 24*day + hour2 - 23
    m[h1.index:h2.index] <- m[h1.index:h2.index] + 1
  }

  # Subset observations where session goes into next day
  df.diffday <- df %>% filter(start_wday != end_wday)
  # Only run the next part if there are multi-day cases
  if ( nrow(df.diffday > 0) ) {
      # populate matrix for obs occurring over span of multiple days
    for (i in 1:nrow(df.diffday)) {
      day1 <- df.diffday$start_wday[i]
      day2 <- df.diffday$end_wday[i]
      hour1 <- df.diffday$start_hour[i]
      hour2 <- df.diffday$end_hour[i]
      # get column numbers for filling in matrix
      h1.index <- 24*day1 - 23 + hour1
      h2.index <- 24*day2 - 23 + hour2
      # first day - up to hour 24 of day1
      m[h1.index:(24*day1)] <- m[h1.index:(24*day1)] + 1
      # second day - begin at hour 1 of day2
      m[(24*day2 - 23):h2.index] <- m[(24*day2 - 23):h2.index] + 1
    }
  }
  return(m)
}
#household.vec(testdf)
#unique(testdf$start_hour)
```

Now we can repeat this for every household to generate a matrix with 24*7 columns
and n_households rows.

```{r}
get_matrix <- function(df) {
  start_time <- Sys.time()
  # strip unneeded columns from dataframe
  machine_ids <- unique(df$machine_id) # the unique machine ids in this sample
  n_machines <- length(unique(df$machine_id)) # num of unique machines/households
  
  # init n x (24*7 + 1) matrix to hold machine_id and
  # activity stats for each day/hour combo
  mat <- matrix(data = 0, nrow = n_machines, ncol = (24*7 + 1))
  # iterate through all the machine_ids and add data to matrix
  # one row per machine_id
  for (i in 1:n_machines) {
    id <- machine_ids[i]
    mat[i,] <- c(household.vec(df[df$machine_id == id,]), id)
    # notify user when halfway done
    if (i == round(n_machines/2)) {
      print("Halfway done...")
    }
  }
  # print time taken to complete
  end_time <- Sys.time()
  print(end_time-start_time)
  
  # return populated 3D array with number of times active at each interval
  return(mat)
}
```

Test our matrix-generating code, then convert to dataframe and rename columns for interpretability.
Columns will be named with the format `DayOfWeek.Hour`.

```{r}
matrix <- get_matrix(df0.time)# yay, it works
#str(act0)
format(object.size(matrix),units="KB")
# create vector of column names denoting day of week + hour of day
days.of.week <- c("Sun", "Mon", "Tue", "Wed", "Thur","Fri","Sat")
hours.of.day <- as.character(paste0(seq(0,23)))
# empty vector for column names
col.names <- c()
# create column names giving day of week and hour of day
for (day in days.of.week) {
  col.names <- append(col.names, paste0(day,".",hours.of.day))
}
# add last colname for machine_id
new.names <- c(col.names,"machine_id")
# now lets apply the better column names
output.df <- data.frame(matrix)
colnames(output.df) <- new.names
```


### Full data matrix generation

```{r}
# extract + format dates and times
df0.cleantimes <- cleantimes(df0)
# generate the matrix of hourly activity patterns
m0 <- get_matrix(df0.cleantimes)

# create vector of column names denoting day of week + hour of day
days.of.week <- c("Sun", "Mon", "Tue", "Wed", "Thur","Fri","Sat")
hours.of.day <- as.character(paste0(seq(0,23)))
# empty vector for column names
col.names <- c()
# create column names giving day of week and hour of day
for (day in days.of.week) {
  col.names <- append(col.names, paste0(day,".",hours.of.day))
}
# add last colname for machine_id
new.names <- c(col.names,"machine_id")
# now lets apply the better column names
output.df <- data.frame(m0)
colnames(output.df) <- new.names
```

Save matrix as RDS file:

```{r}
saveRDS(output.df, file="/Users/sebastiantsai/Desktop/Applied Math Thesis/df0_sample_activity.rds")

saveRDS(array, file="/Users/sebastiantsai/Desktop/Applied Math Thesis/df0_3d_array_unscaled.rds")
##saveRDS(household_ids, file="/Users/sebastiantsai/Desktop/Applied Math Thesis/df0_3d_array_household_ids.rds")
```

First, prepare the dataset by standardizing the variables.

Note: `scale is generic function whose default method centers and/or scales the columns of a numeric matrix.`

### Save to RDS
```{r}
scaled.df <- data.frame(scale(output.df %>% select(!machine_id)))

saveRDS(scaled.df, file="/Users/sebastiantsai/Desktop/Applied Math Thesis/df_sample_activity_scaled.rds")

#readRDS("/Users/meganmcqueen/Box/Senior Project/data/df_sample_activity_scaled.rds")
```

